---
title: "Cyclist"
output: html_document
date: "2022-10-23"
---

# Cyclistic Bike-Sharing Analysis

## Overview

This project is part of Google Data Analytics Course on Coursera.

In 2016, Cyclistic launched a successful bike-sharing offering. Since then, the program has grown to a fleet of 5,824 bikes that are geo-tracked and locked into a network of 692 stations across Chicago. Bikes can be accessed from one station and returned to another at any time in the system.

To date, Cyclistic's marketing strategy has relied on efforts to build general awareness and attract a broad consumer segment. One approach that helps make this happen is price flexibility: one-way tickets, full-day passes, and annual memberships. Customers who purchase a one-way ticket or a full day ticket are referred to as casual riders. Customers who purchase an annual membership are Cyclistic members. Therefore, the marketing analyst team needs to better understand the differences between annual members and regular riders in this month.

## Dataset Preview

For this project, we will use the data for September 2022 [here](https://divvy-tripdata.s3.amazonaws.com/index.html).

This dataset is in the form of CSV (comma-separated value) with 701,339 rows of data. The field in data included:

1.  ride_id
2.  rideable_type
3.  started_at
4.  ended_at
5.  start_station_name
6.  start_station_id
7.  end_station_name
8.  end_station_id
9.  start_lat
10. start_lng
11. end_lat
12. end_lng
13. member_casual

## Steps

1.  Import data.
2.  Clean and add data to prepare for analysis.
3.  Perform a descriptive analysis.

## Import Data

```{r message=FALSE}
library("readr")
library("dplyr")

#Importing data
divvy_202209 <- read_csv("~/divvy-tripdata/202209-divvy-publictripdata.csv")

#Checking data dimension
dim(divvy_202209)

#Checking columns names
colnames(divvy_202209)
```

## Data Cleaning and Preparation

We will analyze the behavior of member and casual user, so we only need ride_id, rideable_type, started_at, ended_at, and member_casual.

```{r message=FALSE}
#Selecting columns
divvy_202209 <- divvy_202209 %>% 
  select(c(ride_id, rideable_type, started_at,ended_at,member_casual))

head(divvy_202209)
```

We will analyze the ride length of users, so we will make new column that shows the length of ride in minutes. We will count the difference between start and end time of ride.

```{r message=FALSE}
#Make new column for ride length
divvy_202209$ride_length <- difftime(divvy_202209$ended_at,divvy_202209$started_at)
divvy_202209$ride_length <- as.double(divvy_202209$ride_length)
```

Then, we remove the row with negative value of ride length.

```{r message=FALSE}
#Remove the row with negative value of ride length
divvy_202209_v2 <- divvy_202209[!divvy_202209$ride_length<0,]
```

We will analyze bike usage for each day, so we will make new columns for started day, ended day, and each day in week.

```{r message=FALSE}
#Make new column for bike usage in start_date, end_date
divvy_202209_v2$started_date <- as.Date(divvy_202209_v2$started_at)
divvy_202209_v2$ended_date <- as.Date(divvy_202209_v2$ended_at)
divvy_202209_v2$Day_of_week <- format(as.Date(divvy_202209_v2$started_date), "%A")
```

We will check data summary of ride length and the boxplot to see if there are any outliers.

```{r message=FALSE}
#Summary of ride length
summary(divvy_202209_v2$ride_length)
```

We get that the minimum ride length is 0 sec, the 1st quantile is 353 secs, the median is 613 secs, the 3rd quantile is 1075 secs, the maximum ride length is 1661854 secs, and the mean is 1160 secs. From the summary, we get the possibilities that the usage of bike can be more than one day. It can affect our analyze if we want to check the average of ride length for each day.

Furthermore, we need to filter the data so we only analyze the ride that start and end at the same day.

```{r message=FALSE}
#Filtering the data
divvy_202209_v3 <- divvy_202209_v2 %>% 
  filter(started_date == ended_date)
```

We check the summary from the new dataset.

```{r message=FALSE}
#Summary of ride length in new dataset
summary(divvy_202209_v3$ride_length)
```

## Analyzing Data

We will compare the mean, minimum value, maximum value, and median of member and casual users in ride length for same-day trip.

```{r message =FALSE}
#Compare member and casual user in ride length
aggregate(divvy_202209_v3$ride_length ~ divvy_202209_v3$member_casual, FUN = mean)
aggregate(divvy_202209_v3$ride_length ~ divvy_202209_v3$member_casual, FUN = min)
aggregate(divvy_202209_v3$ride_length ~ divvy_202209_v3$member_casual, FUN = max)
aggregate(divvy_202209_v3$ride_length ~ divvy_202209_v3$member_casual, FUN = median)
```

From the data, we can see that the mean of ride lengths for same-day trip is 1154.857 secs for casual users and 744.738 for member users. 

The minimum ride length for both users is 0 second and the maximum ride lengths for casual and member users are 77295 secs and 69947 secs, consecutively. 

The median of ride length for casual and member users are 717 secs and 545 secs.

Then, we compare the total of member and casual user based on rideable type.

```{r message=FALSE}
#Compare member and casual user with rideable type
library("janitor")
divvy_202209_v3 %>% 
  tabyl(rideable_type,member_casual) %>% 
  adorn_totals()

```

We can see that there are no member users who use docked bike. The total of member users who use classic bike and electric bike is larger than casual users, and the total of member users is larger than casual users.

Then, we compare the total of member and casual user based on ride length each day.

```{r message=FALSE}
#Compare member and casual user in ride length each day
aggregate(divvy_202209_v3$ride_length ~ divvy_202209_v3$member_casual + divvy_202209_v3$Day_of_week, FUN =mean)
```

The days in the output are not ordered, so we need to order it first.

```{r message=FALSE}
#Ordering day
divvy_202209_v3$Day_of_week <- ordered(divvy_202209_v3$Day_of_week, levels=c("Monday", "Tuesday", "Wednesday", "Thursday","Friday","Saturday","Sunday"))
```

Then we can check the data with new order.

```{r message=FALSE}
#Compare member and casual user in ride length each day
aggregate(divvy_202209_v3$ride_length ~ divvy_202209_v3$member_casual + divvy_202209_v3$Day_of_week, FUN =mean)
```

For make the data easeier to see, we will make the visualization.

```{r message=FALSE}
library("ggplot2")
library("lubridate")

#Analyze number of rides on weekday by user type 
divvy_202209_v3 %>%
  mutate(weekday = lubridate::wday(started_date, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday, rideable_type) %>%  #groups by usertype and weekday'
  summarise(number_of_rides = n(), #calculates the number of rides and average duration
            average_duration = mean(ride_length)) %>% # calculates the average duration
  arrange(member_casual, weekday) %>% # sorts
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + 
  labs(title = "Number of Rides on Weekday", subtitle = "Based on same-day trip in September, 2022", x="Day", y="Number of Rides", fill="User type") +
  theme(plot.title =element_text(hjust = 0.5), plot.subtitle =element_text(hjust = 0.5) ) 
```

From the graphic, we can see that the number of rides for member user is higher than casual users, except on Saturday and Sunday.

The highest number of rides by member users is on Thursday and the highest number of rides by casual users is on Friday and Saturday.

```{r message=FALSE}
#Analyze number of rides on weekday by user type and bike type
divvy_202209_v3 %>%
  mutate(weekday = lubridate::wday(started_date, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday, rideable_type) %>%  #groups by usertype and weekday'
  summarise(number_of_rides = n(), #calculates the number of rides and average duration
            average_duration = mean(ride_length)) %>% # calculates the average duration
  arrange(member_casual, weekday) %>% # sorts
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + 
  labs(title = "Number of Rides on Weekday", subtitle = "Based on same-day trip in September, 2022", x="Day", y="Number of Rides", fill="User type") +
  theme(plot.title =element_text(hjust = 0.5),plot.subtitle = element_text(hjust = 0.5)) +
  facet_wrap("rideable_type",ncol=2,nrow =2)

```

From the graph, we can see the highest number of rides by casual users and member user with classic bike are on Saturday and Thursday, consecutively. 

The highest number of rides by casual users and member user with electric bike are on Friday and Thursday, consecutively. 

The highest number of rides by casual users with docked bike is on Saturday.

```{r message=FALSE}
#Analyze average duration on weekday by user type 
divvy_202209_v3 %>%
  mutate(weekday = lubridate::wday(started_date, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday'
  summarise(number_of_rides = n(), #calculates the number of rides and average duration
            average_duration = mean(ride_length)) %>% # calculates the average duration
  arrange(member_casual, weekday) %>% # sorts
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + 
  labs(title = "Average Duration on Weekday",subtitle = "Based on same-day trip in September", x="Day", y="Average Duration (in secs)", fill="User type") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
```

We can see from the graphic that casual users have more trip average duration on each day than member users, with the highest is on Saturday. The highest trip average duration for member users is on Saturday and Sunday.


```{r message=FALSE}
#Analyze average duration on weekday by user type 
divvy_202209_v3 %>%
  mutate(weekday = lubridate::wday(started_date, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday, rideable_type) %>%  #groups by usertype and weekday'
  summarise(number_of_rides = n(), #calculates the number of rides and average duration
            average_duration = mean(ride_length)) %>% # calculates the average duration
  arrange(member_casual, weekday) %>% # sorts
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + 
  labs(title = "Average Duration on Weekday",subtitle = "Based on same-day trip in September", x="Day", y="Average Duration (in secs)", fill="User type") +
  theme(plot.title =element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  facet_wrap("rideable_type",ncol=2,nrow =2)
```
We can see the trip average durations of docked bike are higher than other type of bikes, with the highest is on Saturday.

For classic bike, the highest trip average durations are by casual users on Saturday and Sunday. Then, the highest trip average durations for classic bike by member users are also on Saturday and Sunday.

For electric bike, the highest trip average durations are by casual users on Saturday. Then, the highest trip average durations for electric bike by member users are on Saturday and Sunday.
